# Week 6. Analysis of longitudinal data

```{r}
date()
```

```{r, message=FALSE}
# Packages
library(tidyverse); library(ggplot2); library(dplyr); library(tidyr); library(lme4); library(patchwork)
```

### 1. PART 1 using RATS data

The data used in the exercise comes from a nutrition study conducted in three groups of rats (16 rats in total). These rats were put on different diets and their weights were recorded repeatedly over a 9-week period.

```{r}
# Read data:
RATSL <- read_csv("data/RATSL.csv")

# Take a look
glimpse(RATSL)
head(RATSL)
names(RATSL)

# Factor variables id and group
RATSL$id <- factor(RATSL$id)
RATSL$group <- factor(RATSL$group)

```

Plot the weights for each rat, differentiating between treatment groups.

```{r}
# Draw the plot
ggplot(RATSL, aes(x = time, y = weight, group = id)) +
  geom_line(aes(linetype = group)) +
  facet_grid(. ~ group, labeller = label_both) +
  theme(legend.position = "none") + 
#  scale_y_continuous(limits = c(min(RATSL$weight), max(RATSL$weight))) +
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10))+
  scale_y_continuous(name = "Weight (grams)")
```

Levels of weight differ but the weight gain trajectories seem quite similar across the three groups. What differs is the starting weight. Especially rats in group 1 are much lighter compared to rats in groups 2 and 3.

Standardize weight and draw the plot again.

```{r}
# Standarize
RATSL <- RATSL %>%
  group_by(time) %>%
  mutate(stdweight = (weight - mean(weight)) / sd(weight)) %>%
  ungroup()

# Plot
ggplot(RATSL, aes(x = time, y = stdweight, group = id, color = group)) +
  geom_line(aes(linetype = group)) +
  facet_grid(. ~ group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10))+
  scale_y_continuous(name = "Standardized weight (grams)")
```

After standardization weights of rats in group 3 seem to somewhat decline.

Summary graph 1: Mean profiles of weight for each of the three groups, with standard errors.

```{r}
RATSL <- RATSL %>%
  group_by(group) %>%
  mutate(count = n_distinct(id)) %>% 
  ungroup()

# Create a summary data with the mean and standard error of weight
RATSS <- RATSL %>%
  group_by(group, time) %>%
  summarise(mean = mean(weight), se = sd(weight)/sqrt(count) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSS)

# Plot the mean values by treatment group
ggplot(RATSS, aes(x = time, y = mean, linetype = group, shape = group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  scale_y_continuous(name = "mean(weight) +/- se(weight)")
```

Summary graph 2 - are there outliers?

```{r}
# Create a summary dataset by treatment and id with the mean of days 8 to 64 as summary measure (ignoring day 1 baseline)
RATSS2 <- RATSL %>%
  filter(time > 1) %>%
  group_by(group, id) %>%
  summarise(mean=mean(weight) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSS2)

# Box plot of means by treatment group
ggplot(RATSS2, aes(x = group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=2, fill = "white") +
  scale_y_continuous(name = "mean(weight), days 8-64")

```

All three groups seem to have one outlier each. Let's remove these.
```{r}
RATSS3 <- RATSS2 %>%
  filter(mean > 250 & group == '1' | 
           mean < 550 & group == '2' | 
           mean > 500 & group == '3')

# Plot again without outliers
ggplot(RATSS3, aes(x = group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=2, fill = "white") +
  scale_y_continuous(name = "mean(weight), days 8-64")
```

Fit an ANOVA model.

```{r}
fit = aov(mean ~ group, data = RATSS3) 
summary(fit)
```
The test confirms what the plots have already shown: the mean weights across the three groups are not the same.

Let's add the baseline from the original data as a new variable to the summary data (now using the data with the outliers) to fit a linear model using the baseline as an additional control.
```{r}
# Original data in wide format
RATS <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt", sep = "", header = T)

# New column
RATSS4 <- RATSS2 %>%
  mutate(baseline = RATS$WD1)

# Fit a linear model with the mean as the response 
fit <- lm(mean ~ baseline + group , data = RATSS4)

# Compute the analysis of variance table for the fitted model with anova()
anova(fit)
```
We see that the baseline weight is strongly related to the weight measurements in the following weeks but there is no evidence of a difference between groups even after conditioning on the baseline value.


### 2. PART 2 using BPRS data

The data used in this exercise corresponds to 40 men. These men were randomly assigned to two treatment groups and each of them was rated on a brief psychiatric rating scale (BPRS). Measurements were done before treatment (week 0) and then weekly for eight weeks.

```{r}
# Read data:
BPRSL <- read_csv("data/BPRSL.csv")

# Take a look
glimpse(BPRSL)
head(BPRSL)
names(BPRSL)

# Factor variables treatment and subject
BPRSL$treatment <- factor(BPRSL$treatment)
BPRSL$subject <- factor(BPRSL$subject)

str(BPRSL)
```
As a first step lets draw a plot identifying the observations in each treatment group but ignoring the longitudinal nature of the data.

```{r}
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject, color = treatment)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)), name = "BPRS") +
  scale_x_continuous(name = "Week", breaks = seq(0, 8, 1))
```

BPRS values show a somewhat declining trend over time but the plot suggests there is little difference between the two treatment groups.

Next a linear model is fitted, still ignoring the repeated-measures structure of the weights in the data.

```{r}
# Create a regression model BPRS_reg
BPRS_reg <- lm(bprs ~ week + treatment, data = BPRSL)

# print out a summary of the model
summary(BPRS_reg)
```

There seems to be indeed a decreasing trend over the weeks but the model gives no evidence for a significant difference between treatment groups.

Next, a random intercept model is fitted which allows the intercepts of individual men to differ from other men.

```{r}
# Create a random intercept model
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)

# Print the summary of the model
summary(BPRS_ref)
```

Now let's try letting also the slopes differ, with a random intercept and random slope model.

```{r}
# Create a random intercept and random slope model
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)

# Print a summary of the model
summary(BPRS_ref1)
```

Week remains significant but no difference appears between the two treatment groups.

Compare the two previous models:

```{r}
anova(BPRS_ref1, BPRS_ref)
```

Random slopes and intercept model performs slightly better. Let's add an interaction between week and treatment group.

```{r}
BPRS_ref2 <- lmer(bprs ~ week * treatment + (week | subject), data = BPRSL, REML = FALSE)

# Print a summary of the model
summary(BPRS_ref2)

# perform an ANOVA test on the two models
anova(BPRS_ref2, BPRS_ref1)
```

The interaction is not significant and the ANOVA test suggests to stay with the previous ref1 model. I use that for plotting.

```{r}
# Plot the observed values
p1 <- ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject, color = treatment)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)), name = "Observed bprs") +
  scale_x_continuous(name = "Week", breaks = seq(0, 8, 1))

# Create a vector of the fitted values
fitted <- fitted(BPRS_ref1)

# Add a new column with fitted values BPRSL
BPRSL <-  BPRSL %>% 
  mutate(fitted = fitted)
head(BPRSL)

# Plot the fitted values of bprs
p2 <- ggplot(BPRSL, aes(x = week, y = fitted, linetype = subject, color = treatment)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)), name = "Fitted bprs") +
  scale_x_continuous(name = "Week", breaks = seq(0, 8, 1))

p1 | p2
```

Treatment groups seem very alike.
